---
title: "Introduction to dcmR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to dcmR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
```

# Overview

`dcmR` is an R port of the Dynamic Causal Modelling (DCM) routines from
the MATLAB SPM12 toolbox. It estimates effective connectivity among
brain regions from fMRI BOLD time series using variational Laplace
inversion (Friston et al., 2003).

This vignette walks through a minimal end-to-end example using the
synthetic `toy_dcm` dataset shipped with the package.

```{r setup}
library(dcmR)
```

# A two-region toy model

The package includes a synthetic 2-region, 64-sample DCM whose only
purpose is to exercise the inversion machinery quickly.

```{r}
data(toy_dcm)
str(toy_dcm, max.level = 1)
```

The structure follows the SPM12 convention:

- `a`, `b`, `c`, `d` are the connectivity / modulatory / driving /
  non-linear adjacency arrays.
- `Y$y` is the BOLD data (rows = scans, columns = regions).
- `U$u` is the input design at the microtime resolution.
- `options` controls the model variant (deterministic, two-state, etc.).

# Inverting the model

```{r, eval = FALSE}
fit <- dcm_estimate(toy_dcm)
```

`dcm_estimate` returns the input DCM augmented with posterior
fields:

- `Ep` — posterior expectation of the parameters (same shape as the
  prior).
- `Cp` — posterior covariance.
- `F`  — variational free energy (the log-evidence proxy used for
  model comparison).
- `y`, `R` — predicted BOLD and residuals.

```{r, eval = FALSE}
round(fit$Ep$A, 3)   # estimated connectivity
fit$F                # log-evidence
```

By default progress messages are emitted via `message()`; suppress
them with `suppressMessages()` if you are scripting.

# Model comparison

For nested model comparison (Bayesian model reduction), use
`dcm_log_evidence` with the original posterior and the reduced prior.
For an approximate AIC / BIC summary of a single fit, use
`dcm_evidence`.

```{r, eval = FALSE}
ev <- dcm_evidence(fit)
ev$aic_overall
ev$bic_overall
```

# Tuning

The finite-difference step used by the numerical Jacobian is exposed
via `dcm_options()`:

```{r}
dcm_options()$GLOBAL_DX
```

The default `exp(-8)` is what SPM12 uses; widen it (e.g. `1e-4`) if
you observe numerical issues with very flat regions of the likelihood.

# Acknowledgment

`dcmR` is a derivative work of the MATLAB SPM12 toolbox, distributed
under GPL-2 by the Wellcome Centre for Human Neuroimaging. See the
`LICENSE.note` file in the package source for the list of ported
routines.

# References

- Friston, K.J., Harrison, L., Penny, W. (2003). Dynamic causal
  modelling. *NeuroImage*, **19**(4), 1273-1302.
- Friston, K.J., Mattout, J., Trujillo-Barreto, N., Ashburner, J.,
  Penny, W. (2007). Variational free energy and the Laplace
  approximation. *NeuroImage*, **34**(1), 220-234.
